---
title: "Standard Deviation Transition Temperature (Tt) Analysis with batch_plot"
author: "David Tyrpak"
output:
  pdf_document:
    fig_caption: no
    fig_crop: no
    fig_height: 8
    fig_width: 9
  html_notebook:
    fig_caption: no
    fig_crop: no
    fig_height: 10
    fig_width: 12
  html_document:
    df_print: paged
geometry: margin=1cm

---


# INTRODUCTION
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This notebook highlights the standard deviation approach to 
determing intracellular transition temperature

Let's begin by loading and cleaning the data.

## Loading data
We will load StandardDevData.txt which contains pixel standard deviation values for the GFP-V60 cells.
Each row in this tab delimited txt file contains the pixel standard deviation values
for that particular cell at that particular temperature.
```{r}
example.df <- read.table(file.choose(), header = T) #choose StandardDevData.txt
```

## The batch_plot function

This function will calculate the Tt for each cell in our data frame (after we organize the data).
These are the arguments of the function:

list_of_lists: list class; this is a list containing lists of row indexes for individual cells.
We will create this data structure further below.
  
data_frame: data frame class; the data frame that contains your temperature and standard deviation values.
In our case, the data_frame is example.df
  
threshold_limit: int class; how many standard deviations away from the baseline you want your threshold to be. 
  
baseline_end: int class; the number of data points you want to include 
in your calculation of the baseline standard deviation (e.g. "baseline_end = 20" means the 
first 20 points of the data set are used in the calculation of the mean and standard deviation)

temp.column.index: int class; the column number that contains your temperature data

standard.dev.column.index: int class; the column number that contains your pixel standard deviation data

```{r}
batch_plot <- function(list_of_lists, data_frame, threshold_limit = 3, baseline_end = 20, temp.column.index, standard.dev.column.index){
  
  
  
  indexes <- lapply(list_of_lists, function(x) {c(start  = x[1], stop = tail(x, n = 1))}) ##retrieve first and last index for that well/cell
  
  i = 1
  Tts <- vector(mode = "numeric", length = length(list_of_lists)) # will store the Tts for each cell
  
  ##for loop to determine Tt and plot a scatter plot for each cell in list_of_lists
  for (index in indexes) {
    start = index[1]
    stop = index[2]
    
    temps <- data_frame[ ,temp.column.index][start:stop] ##Celsius data points for this cell
    
    sds <- data_frame[,standard.dev.column.index][start:stop] ##corresponding standard deviation points for this cell
    
    mean <- mean(sds[1:baseline_end]) ##mean of the datapoints specified by baseline_end
    
    std_dv <- sd(sds[1:baseline_end]) ##calculate the standard deviation around these points
    
    std_normals <- (sds - mean)/std_dv ## convert the standard deviation pixel values to a standard normal distribution
    
    #threshold <- threshold_limit*std_dv ##threshold is threshold_limit times the standard deviation from the baseline/mean
    
    Tt <- which(std_normals > threshold_limit)[1]  ## first data point which crosses the threshold
    
    Tts[i] <- temps[Tt] ##update Tts vector to contain the Tt for this cell
    
    
    #if you want to save each plot, you can call the tiff function. 
    #tiff(filename = paste0(names(indexes)[i], ".tiff"), width  = 7, height = 5, units = "in", res = 300)
    
     par(mar=c(5,6,4,1)+.1) #adjust maregins so that axis titles arent cutoff
     
    plot(temps, std_normals, xlab = "Celsius", ylab = "Z-Score", main = names(indexes)[i], ylim = c(-10, 10), cex.main=2, cex.lab=2, cex.axis=2)
    abline(h = 0) ##mean of first 20 points will be zero because we converted to standard normal
    abline(v = Tts[i], col = "red", pch = 19)
    abline(h = 0 + threshold_limit, lty = 4, col = "red") ## 
    # abline(h = 0 - threshold_limit, lty = 4, col = "red") ##optional line beloiew the baseline
    
    #dev.off() only need this if you want to automatically save with tiff() function
    i = i +1
    
    
  }
  names(Tts) <- names(indexes)
  Tts
}

```

### Organizing Data for batch_plot

We need to extract the row numbers (aka row indices) for each
unique cell in our example.df data.frame object.
We can use the cellID column, as it uniquely identifies each cell.
```{r}

indices <- sapply(unique(example.df$cellID), function(x) which(example.df$cellID == x)); indices
names(indices) <- unique(example.df$cellID)
indices

```

### Running batch_plot

Now we can use the batch_plot function and examine different numbers of 
baseline data points and thresholds.
```{r}
# default values
Tts.a <- batch_plot(GFPV60.indices, GFPV60.df, baseline_end = 20, threshold_limit = 3, temp.column.index = 4, 
                  standard.dev.column.index = 5)

mean(na.omit(Tts.a)) # get the mean, after removing NA cells (i.e. cells which didnt pass the threshold)

# default baseline and threshold of 5
Tts.b <- batch_plot(GFPV60.indices, GFPV60.df, baseline_end = 20, threshold_limit = 5, temp.column.index = 4, 
                  standard.dev.column.index = 5)

mean(na.omit(Tts.b))


#change baseline number to 10 and keep threshold of 5
Tts.c <- batch_plot(GFPV60.indices, GFPV60.df, baseline_end = 10, threshold_limit = 5, temp.column.index = 4, 
                  standard.dev.column.index = 5)

mean(na.omit(Tts.c))



```

